---
title: "137 Project"
author: "Quynh Trinh,"
date: "2025-05-30"
output:
  pdf_document: default
  html_document: default
---

```{r}
# Load required libraries
library(tidyverse)      # For data manipulation and visualization
library(forecast)       # For Box-Cox transformation and time series analysis
library(tseries)        # For additional time series functions
library(ggplot2)        # For plotting
library(carData)
```


# Time Series Plot for the Data
```{r}
load("finalproject.Rdata")
ts.plot(finalPro_data)
```

Quick look:

- Upward trend
- Some fluctuations and short-term volatility
- Non-stationary



# GDP Model
## Summary GDP

```{r}
# 1. Diagnose model for GDP
model_gdp <- lm(GDP ~ Year + Growth + CPI + Imports + Exports + Population, data = finalPro_data)
summary(model_gdp)

```
- Significant variables: Year, Exports

## Residuals Diagnostic

### ACF and Ljung-Box

```{r}
#ACF GDP
acf(residuals(model_gdp), main="ACF of Residuals GDP")

# L-jung-box
Box.test(residuals(model_gdp), lag = 20, type = "Ljung-Box")

```
Quick look:

- Autocorrelation remains positive before approaching zero
- The pattern shows strong autocorrelation and persistence in the residuals
- Slow decay --> non-stationary or trending series


## Plot GDP

```{r}
# Diagnostic plots for GDP model
par(mfrow = c(2,2))
plot(model_gdp)
```
Quick interpretation:

- Residuals vs fitted: looks good, no obvious pattern suggests linearity
- QQ plot: residuals are close to the line, suggests approximately normally distributed
- Scale-Location: slightly upward trend, meaning the variance of errors is not constant
- Residuals vs Leverage: most points have low leverage so this is good

Recommendations:

- Transforming the dependent variable; considering Box-Cox or log transformation


## Box-Cox for GDP (Draft)

```{r}
library(MASS)
# Perform Box-Cox transformation
bc <- boxcox(model_gdp, lambda = seq(-2, 2, by = 0.1))

# Find the lambda with the highest log-likelihood
best_lambda <- bc$x[which.max(bc$y)]
cat("Best lambda:", best_lambda, "\n")

# Refit using the Box-Cox transformed GDP:
finalPro_data$GDP_boxcox <- (finalPro_data$GDP^best_lambda - 1) / best_lambda
model_gdp_boxcox <- lm(GDP_boxcox ~ Year + Growth + CPI + Imports + Exports + Population, data = finalPro_data)
summary(model_gdp_boxcox)

par(mfrow = c(2,2))
plot(model_gdp_boxcox)
```

## (Draft) Use auto.arima()

```{r}
# GDP time series covers years 1960–2017:
GDP_ts <- ts(finalPro_data$GDP, start = 1960, frequency = 1)

# Use auto.arima to find the best ARIMA model
best_arima_gdp <- auto.arima(GDP_ts)

# Show model summary
summary(best_arima_gdp)

# If you want to plot diagnostics:
checkresiduals(best_arima_gdp)
```
Summary:
- ARIMA(0,1,0) model fits the GDP time series pretty well
- Residuals show no significant autocorrelation
- Errors appear approximately normal
- Good baseline so far

# Imports Model

### Summary Imports

```{r}
# 2. Diagnose model for Imports
model_imports <- lm(Imports ~ Year + GDP + Growth + CPI + Exports + Population, data = finalPro_data)
summary(model_imports)
```
- Significant variables: Year, CPI, Exports, Population

## Residuals Diagnostics 

```{r}
#ACF Imports
acf(residuals(model_imports), main="ACF of Residuals Imports")

# L-jung-box
Box.test(residuals(model_imports), lag = 20, type = "Ljung-Box")

```
Quick look:

- ACF plot shows no significant autocorrelation at higher lags
- Possible AR(1) model
- Ljung-box p-value > 0.1, so residuals are likely white noise

### Plot Imports

```{r}
# Diagnostic plots for Imports model
par(mfrow = c(2,2))
plot(model_imports)

```

Quick interpretation:

- Residuals vs fitted: linearity assumption is reasonably met
- QQ plot: approximately normally distributed
- Scale-Location: variance of the residuals is roughly constant
- Residuals vs Leverage: most points have low leverage

## (Draft) auto.arima() for Imports

```{r}
# Imports time series covers years 1960–2017:
Imports_ts <- ts(finalPro_data$Imports, start = 1960, frequency = 1)

# Use auto.arima to find the best ARIMA model
best_arima_imports <- auto.arima(Imports_ts)

# Show model summary
summary(best_arima_imports)

# Plot diagnostics:
checkresiduals(best_arima_imports)
```
Summary:

- ARIMA(0,1,2) model fits the Imports time series well
- Residual diagnostics shows good model fit with no significant autocorrelation left
